---
title: "STAT321/421"
author: "Gregory J. Matthews"
format: revealjs
editor: visual
execute: 
  echo: true
---

# Data viz
 
 - We are going to use ggplot
 - ggplot works with data frames
 - We add layers

## Histogram
```{r}
library(ggplot2)
library(palmerpenguins)

ggplot(aes(x = bill_length_mm),data = penguins) + geom_histogram()

```

## Kernel Density Estimators
```{r}
library(ggplot2)
library(palmerpenguins)

ggplot(aes(x = bill_length_mm),data = penguins) + geom_density()

```

## Boxplot 

```{r}
ggplot(aes(x = bill_length_mm),data = penguins) + geom_boxplot()
```

## Barplot 

```{r}
ggplot(aes(x = species),data = penguins) + 
  geom_bar(fill= "red")
```

## Barplot 

```{r}
ggplot(aes(x = species, color = species),data = penguins) + geom_bar()
```

## Barplot 

```{r}
ggplot(aes(x = species, fill = species),data = penguins) + geom_bar()
```

## Side-by-side boxplots
```{r}
ggplot(aes(x = species, y = bill_length_mm), data = penguins) + 
geom_boxplot()
```

## Scatterplot 

```{r}
ggplot(aes(x = bill_length_mm, y = bill_depth_mm),data = penguins) + 
  geom_point()
```

## Scatterplot 

```{r}
ggplot(aes(x = bill_length_mm, y = bill_depth_mm),data = penguins) + 
  geom_point(color = "salmon")
```

## Scatterplot 

```{r}
ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species), data = penguins) +
  geom_point()
```

## Scatterplot 

```{r}
ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species), data = penguins) + 
  geom_point() + 
  geom_smooth()
```

## Faceting

```{r}
ggplot(aes(x = bill_length_mm, y = bill_depth_mm),data = penguins) + 
  geom_point() + 
  geom_smooth() + 
  facet_grid(~species)
```

## Faceting

```{r}
ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = island),data = penguins) + 
  geom_point() + 
  geom_smooth() + 
  facet_grid(~species)
```

## Faceting

```{r}
ggplot(aes(x = bill_length_mm, y = bill_depth_mm), data = penguins) +
  geom_point() +
  geom_smooth() +
  facet_grid(species ~ 1)
```

## Faceting

```{r}
ggplot(aes(x = bill_length_mm, y = bill_depth_mm), data = penguins) +
  geom_point() +
  geom_smooth() +
  facet_grid(species ~ island)
```


## Heat maps

 - You're going to want this for your simulation studies at the end of the semester.   
 
```{r}
df <- expand.grid(x = 0:5, y = 0:5)
set.seed(1)
df$z <- runif(nrow(df))
ggplot(df, aes(x, y, fill = z)) +
geom_tile() + coord_fixed()

#This also works
#ggplot(df, aes(x, y, fill = z)) +
#geom_raster()
```


## Heat Map

```{r}
#| echo: false
df <- expand.grid(x = 0:5, y = 0:5)
set.seed(1)
df$z <- runif(nrow(df))
ggplot(df, aes(x, y, fill = z)) +
geom_tile() + coord_fixed()
```

## Heat maps: color schemes
 
```{r}
df <- expand.grid(x = 0:5, y = 0:5)
set.seed(1)
df$z <- runif(nrow(df))
ggplot(df, aes(x, y, fill = z)) +
geom_tile() + 
  coord_fixed() + 
  scale_fill_continuous(type = "viridis")
  

#This also works
#ggplot(df, aes(x, y, fill = z)) +
#geom_raster()
```

## Heat maps: color schemes
 
```{r}
#| echo: false
df <- expand.grid(x = 0:5, y = 0:5)
set.seed(1)
df$z <- runif(nrow(df))
ggplot(df, aes(x, y, fill = z)) +
geom_tile() + 
  coord_fixed() + 
  scale_fill_continuous(type = "viridis")
  

#This also works
#ggplot(df, aes(x, y, fill = z)) +
#geom_raster()
```

## Heat maps: custom color schemes
 
```{r}
df <- expand.grid(x = 0:5, y = 0:5)
set.seed(1)
df$z <- runif(nrow(df))
ggplot(df, aes(x, y, fill = z)) +
geom_tile() + 
  coord_fixed() + 
  scale_fill_gradient(low = "#eaaa00", high = "#5a0722")
  

#This also works
#ggplot(df, aes(x, y, fill = z)) +
#geom_raster()
```

## Heat maps: custom color schemes
 
```{r}
#| echo: false
df <- expand.grid(x = 0:5, y = 0:5)
set.seed(1)
df$z <- runif(nrow(df))
ggplot(df, aes(x, y, fill = z)) +
geom_tile() + 
  coord_fixed() + 
  scale_fill_gradient(low = "#eaaa00", high = "#5a0722")
  

#This also works
#ggplot(df, aes(x, y, fill = z)) +
#geom_raster()
```