---
title: "STAT321/421"
author: "Gregory J. Matthews"
format: 
  revealjs:
    chalkboard: true
    slide-number: c/t
    code-line-numbers: false
    linestretch: 1.25
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
editor: visual
execute: 
  echo: true
---

## Rootfinding
 - Newton-Raphson method
 - secant method
 - bisection method
 
## Recall
 - A ***root*** is a solution to the equation $f(x) = 0$.  
 - i.e. Where doe sthe curve cross the $x$-axis? 
 - How is this useful to us in statistics? 
 - Least squares estimation (LSE) is simply a root finding problem.  
 
## Newton-Raphson
 - Consider a function $f$ that is differentiable with continuous derivative $f'$ and a root at $a$.  
 - Let $x_0 \in \mathbb{R}$ and $x_0$ is the best guess of $a$.
 - Now consider a line through the point $(x_0, f(x_0))$ and slope $f'(x_0)$.  This is the best straight line approximation of $f(x_0)$.
 - The equation of this line is $f'(x_0) = \frac{f(x_0)-y}{x_0-x}$. 
 
## Newton-Raphson
 - This straight line crosses the x-axis at a point, say $x_1$.  This should be a better guess for $a$ than $x_0$.  
 - To find $x_1$, let $y=0$ and $x=x_1$.  This yields $x_1 = x_0 - \frac{f(x_0)}{f'(x_0)}$. 
 - Generally:

$$
x_{n+1} = x_{n} - \frac{f(x_n)}{f'(x_n)}
$$

## Newton-Raphson
::: {.smaller}

 - If $f$ is "well-behaved" at $a$ (i.e. $f'(a) \ne 0$ and $f''$ is finite and continuous at $a$) then $x_n$ will quickly converge to $a$.  
 - If $x_n \rightarrow a$, and since $f$ and $f'$ are continuous:
 
$$
a = \lim_{n\rightarrow \infty} x_{n+1} = \lim_{n\rightarrow \infty} \left (x_n - \frac{f(x_n)}{f'(x_n)}\right)
$$


$$
 = \lim_{n\rightarrow \infty} x_n - \frac{f\left(\underset{n \rightarrow \infty}{\lim} x_n\right)}{f'\left(\underset{n \rightarrow \infty}{\lim} x_n\right)} = a - \frac{f(a)}{f'(a)}
$$
:::

## Newton-Raphson
 - In order for this to be true, $f(a) = 0$ (as long as $f'(a) \ne \pm \infty$) which is a root. A good stopping criteria is when $|f(x_n)| < \epsilon$ for some $\epsilon$.

## Newton-Rapson: Example
```{r}
f <- function(x){x^3 + x^2 + x - 1}
fprime <- function(x){3*x^2 + 2*x + 1}  

xvec <- c()
#Initialize vector
xvec[1] <- 0

for (i in 2:10){
xvec[i] <- xvec[i-1] - f(xvec[i-1])/fprime(xvec[i-1])
}
xvec
```
## Newton-Rapson: Different starting values
```{r}
#| echo: false
niter <- 12
d <- data.frame()
for (q in seq(-10,10,3)){
xvec <- c()
#Initialize vector
xvec[1] <- q

for (i in 2:niter){
xvec[i] <- xvec[i-1] - f(xvec[i-1])/fprime(xvec[i-1])
}
d <- rbind(d,data.frame(x = 1:niter,y = xvec, q = q))
}
library(tidyverse)

ggplot(aes(x = x, y = y, color = factor(q)), data = d) + geom_path(show.legend = FALSE) + geom_point(show.legend = FALSE) 

```

## Newton-Rapson: Different starting values
```{r}
#| echo: false
niter <- 12
d <- data.frame()
for (q in seq(-100,100,11)){
xvec <- c()
#Initialize vector
xvec[1] <- q

for (i in 2:niter){
xvec[i] <- xvec[i-1] - f(xvec[i-1])/fprime(xvec[i-1])
}
d <- rbind(d,data.frame(x = 1:niter,y = xvec, q = q))
}
library(tidyverse)

ggplot(aes(x = x, y = y, color = factor(q)), data = d) + geom_path(show.legend = FALSE) + geom_point(show.legend = FALSE) 

```


## Newton Raphson in R: Uniroot
 - Uses Newton-Raphson
 - https://rpubs.com/aaronsc32/newton-raphson-method
 
## BOBYQA
 - Works great in lme4.  
```{r}
library(nloptr)
bobyqa
```
 
 
