---
title: "STAT321/421"
author: "Gregory J. Matthews"
format: revealjs
editor: visual
---

# Useful Links

-   [Installing R and RStudio](https://36-sure.github.io/r-setup.html)

-   [Setting up Git and Github](https://36-sure.github.io/git-setup.html)

-   [Quarto](https://r4ds.hadley.nz/quarto)

-   [Piazza Course Site](https://piazza.com/class/lzyfs2usjv2e6)

-   [Dr. Matthews' Zoom link](https://luc.zoom.us/j/6622906748)

-   [Advanced R](https://adv-r.hadley.nz/)

-   [R for Data Science](https://r4ds.had.co.nz/)

## R Data types and operators

-   Tensors (Vectors, Matrices, Arrays)
-   Lists
-   Data Frames

## Vectors

-   Basic unit of R
-   All elements must be same type
-   Types of vectors:
    -   Numeric
    -   Integer
    -   Character
    -   Logical

## Examples

```{r}
#| echo: true
v1 <- c(413,508,603,860,312)
class(v1)

v2 <- as.integer(c(413,508,603,860,312))
class(v2)

v3 <- c("rock","guts","agrocrag","trash", "spritecoin", "hai davai")
class(v3)

v4 <- c(TRUE, FALSE, TRUE)
class(v4)

v5 <- as.factor(c(103,112,119,125,13))
class(v5)
```

## Operations

All operations are done element wise

```{r}
#| echo: true
#addition
v1 + v2

#substraction
v1 - v2

#Multiplication
v1 * v2

#Division
v1 / v2
```

## Operations

```{r}
#| echo: true
sum(v1)
```

## Subsetting a vector

```{r}
#| echo: true
#pull out a single element
v1[1]

#pull out consecutive elements
v1[2:4]

#remove an element
v1[-2]

#remove consecutive elements
v1[-c(3:4)]

#pull out nonconsecutive elements
v1[c(1,3,5)]
```

## Matrices

## Arrays

## Lists

## Data Frames
